pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-02-27 bmc_article.tex

\author[
   addressref={aff1,aff2},          % id's of addresses, e.g. {aff1,aff2}
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1},
    email={anthony.sonrel@uzh.ch}
]{\fnm{Anthony} \snm{Sonrel}}
\author[
   addressref={aff1},
   corref={aff1},  % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}

@@

\address[id=aff1]{
  \orgname{Department of Molecular Life Sciences and SIB Swiss Institute of Bioinformatics, University of Zürich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics, and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection (suggesting a new R package, \textit{scDblFinder}), normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, permit extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

@@

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can inform analytical decisions (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, HeiserDimred2019, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution (\citealp{viethSystematic2019, CobosDeconvolution2020}). It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline, and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI, \citep{HubertARI1985}) for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \citep{steinley_properties_2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \citep{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown, and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive, and might not always capture improvements in earlier steps of the pipeline. We therefore favour monitoring several complementary metrics across multiple steps of the process.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept of number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies, and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \citep{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \citep{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \citep{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \citep{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals), whose identification is generally not the primary aim of doublet detection methods, since these can be considered homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types) and can be identified through other means.

We tested \textit{DoubletFinder} \citep{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \citep{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \citep{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C). We therefore tested whether this method also improved the accuracy of the clustering across all benchmark datasets, and found that it did even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \citep{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties, for instance through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two, and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to eliminate are cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation, and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g. library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here, and we therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \citep{PapadimitriouHu1998}, and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious, and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

Mitochondrial reads have been associated with cell degradation, and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \citep{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \citep{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \citep{hafemeisterSCtransform2019}, and normalization based on stable genes \citep{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves something analogous to scaling). 
\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq\citep{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \citep{hafemeisterSCtransform2019}. We therefore first assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We next investigated the impact of normalization on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \citep{RousseeuwSil1987} and mutual information. We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones, and was generally but not always beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods, and even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

>>>>

\author[
   addressref={aff1,aff2,aff3},          % id's of addresses, e.g. {aff1,aff2}
   email={pierre-luc.germain@hest.ethz.ch}   % email address
]{\inits{PLG}\fnm{Pierre-Luc} \snm{Germain}}
\author[
   addressref={aff1,aff2},
    email={anthony.sonrel@uzh.ch}
]{\fnm{Anthony} \snm{Sonrel}}
\author[
   addressref={aff1,aff2},
   corref={aff1},  % id of corresponding address, if any
   email={mark.robinson@imls.uzh.ch}
]{\inits{MR}\fnm{Mark D.} \snm{Robinson}}

@@

\address[id=aff1]{
  \orgname{Department of Molecular Life Sciences, University of Zürich}, % university, etc
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff2]{
  \orgname{SIB Swiss Institute of Bioinformatics}, % university, etc
  \city{Zürich},
  \cny{Switzerland}
}
\address[id=aff3]{
  \orgname{D-HEST Institute for Neurosciences, ETH Zürich},
  \street{Winterthurerstrasse 190},
  \postcode{8057}
  \city{Zürich},
  \cny{Switzerland}
}

@@

\begin{abstract} % abstract
%\parttitle{Background} %if any
The massive growth of single-cell RNA-sequencing (scRNAseq) and methods for its analysis still lacks sufficient and up-to-date benchmarks that would guide analytical choices. Moreover, current studies are often focused on isolated steps of the process. Here, we present a flexible R framework for pipeline comparison with multi-level evaluation metrics and apply it to the benchmark of scRNAseq analysis pipelines using datasets with known cell identities. We evaluate common steps of such analyses, including filtering, doublet detection, normalization, feature selection, denoising, dimensionality reduction and clustering. On the basis of these analyses, we make a number of concrete recommendations about analysis choices. The evaluation framework, \textit{pipeComp}, has been implemented so as to easily integrate any other step or tool, allowing extensible benchmarks and easy application to other fields (\url{https://github.com/plger/pipeComp}).

@@

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

Here, we develop a flexible R framework for pipeline comparison and evaluate the various steps of analysis leading from an initial count matrix to a cluster assignment, which are critical in a wide range of applications. We collected real datasets of known cell composition (Table \ref{tab:table1}) and used a variety of evaluation metrics to investigate in a multilevel fashion the impact of various parameters and variations around a core scRNAseq pipeline. Although we use some datasets based on other protocols, our focus is especially on droplet-based datasets that do not include exogenous control RNA (i.e. spike-ins); see Table \ref{tab:table1} and Figure \ref{fig:figure1} for more details. In addition to previously-used benchmark datasets with true cell labels \cite{duoClustering2018,tianMixology2018}, we simulated two datasets with a hierarchical subpopulation structure based on real 10x human and mouse data using \textit{muscat} \cite{CrowellMuscat2019}. 
Since graph-based clustering \cite{satijaSeurat2015} was previously shown to consistently perform well across several datasets \cite{duoClustering2018,tianMixology2018}, we used the \textit{Seurat} pipeline as the starting point to perform an integrated investigation of: 1) doublet identification, 2) cell filtering, 3) normalization, 4) feature selection, 5) dimension reduction, 6) clustering. We compared competing approaches and also explored more fine-grained parameters and variations on common methods. Importantly, the success of methods at a certain analytical step might be dependent on choices at other steps. Therefore, instead of evaluating each step in isolation, we developed a general framework for evaluating nested variations on a pipeline and suggest a multilevel panel of metrics. Finally, we evaluate several recent methods and provide concrete recommendations.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. We therefore favour monitoring several complementary metrics across multiple steps of the process.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept of number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).

While most methods accurately identified the doublets in the 3 cell lines dataset (mixology10x3cl), the other two datasets proved more difficult (Figure \ref{fig:figure3}A). \textit{scDblFinder} achieved comparable or better accuracy than top alternatives while being the fastest method (Figure \ref{fig:figure3}B). Across datasets, cells called as doublets tended to be classified in the wrong cluster more often than other cells (Figure \ref{fig:figure3}C). We also found that \textit{scDblFinder} improved the accuracy of the clustering across all benchmark datasets even when, by design, the data contained no heterotypic doublet (Figure \ref{fig:figure4}).

\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We next investigated the impact of different normalization strategies. Beside the standard log-normalization included in \textit{Seurat}, we tested \textit{scran}'s pooling-based normalization \cite{lunPooling2016}, \textit{sctransform}'s variance-stabilizing transformation \cite{hafemeisterSCtransform2019}, and normalization based on stable genes \cite{linStableGenes2018, deekeStablyExpressed2018}. 
In addition to log-normalization, the standard \textit{Seurat} clustering pipeline performs per-feature unit-variance scaling so that the PCA is not too strongly dominated by highly-expressed features. We therefore included versions of the different normalization procedures, with or without a subsequent scaling (\textit{sctransform}'s variance-stabilizing transformation involves an approach analogous to scaling). 
\textit{Seurat}'s scaling function also includes the option to regress out the effect of certain covariates. We tested its performance by using the proportion of mitochondrial counts and the number of detected features as covariates. Finally, it has been proposed that the use of stable genes, in particular cytosolic ribosomal genes, can be used to normalize scRNAseq \cite{deekeStablyExpressed2018}. We therefore evaluated a simple linear normalization based on the sum of these genes, as well as nuclear genes.

An important motivation for the development of \textit{sctransform} was the observation that, even after normalization, the first principal components of various datasets tended to correlate with library size, suggesting an inadequate normalization \cite{hafemeisterSCtransform2019}. However, as library size tends to vary across subpopulations, part of this effect can simply reflect biological differences. We therefore assessed to what extent the first principal component still retained a correlation with the library size and the number of detected features, removing the confounding covariation with the subpopulations (Figure \ref{fig:figure5}A). The simple step of scaling tended to remove much of the correlation with these features and, in the absence of scaling, standard \textit{Seurat} normalization resulted in fairly high correlation with technical covariates. \textit{sctransform} led to the lowest correlation but most methods, including normalization based on stable genes, were able to remove most of the correlation when combined with scaling. The only exception is \textit{Seurat} normalization with scaling regressing out covariates, which surprisingly increased the correlation in 8 of the 9 datasets.

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure TBD) that performed poorly on Zhengmix and simMix1 datasets in view of the subpopulation silhouette and to a relative poor ARI. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only. Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}
